From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-20 Pi+/paper/finite.tex
Outline the coxeter rewriting system

When swapping two pairs of elements in positions far away, it does not matter in
which order the swapping happens, that is, we can slide the wires.

@@

@@

@@

This construction is called a Coxeter presentation. Writing it formally, we
get the following definition
Definition: Coxeter presentation.

13. There are two main considerations to this choice:
13a. How well the presentation matches the syntactic side of Pi? This is,
how closely does the operational semantics built from these particular
primitive operations match the natural operational semantics of Pi?
13b. How convenient is it to turn the presentation it into rewriting system
and solve the word problem for it?

Thus, the argument:
14. TODO I don't know what to write here, we have to think.

15a. The equations that define Coxeter presentation have a nice property:
as it is written, the LHS is smaller (lexicographically) than the RHS.
Thus, if we could direct them from right to left, it would have a
termination property out of the box.
15b. However, it is not at all obvious that the system would be confluent.
There are many possible critical pairs, such as
TODO diagram 1212 (braid, braid)
or
TODO diagram 3233 (braid, cancel)
As it turns out, there are instances of critical pairs that do not resolve.
A simplest example is the following:
TODO diagram 3213

16. The equations have to be changed. We thus propose a new presentation.
We keep the (cancel) and (swap) equations, but replace (braid) with a more
general:
TODO diagram:

Definition: Long Coxeter presentation.

17. In return for increased complexity of the equations, it satisfies the
diamond property (if we direct the equation). Thus, together with the
(preserved) termination, it creates a confluent rewriting system.
Proof: All the critical pairs resolve, in the appendix.
Definition: A normal form is a list that cannot be reduced any further.

@@

\subsection{Symmetric groups}~\label{subsec:symmetric}

There is an obvious group structure on $\Aut[\Fin[n]]$ given by identity,
composition, and inverse. This is the symmetric group $S_n$ on $n$ symbols. In
the rest of the section we will construct a convenient presentation of this
group.

\vc{this is just a rough draft for now}

\todo{Reference T-algebra presentations as coequalisers (Mac Lane 6.7)}

First, we formally define a presentation of a group.

\todo{Free 1-group, needs to be truncated}

\subsubsection{Coxeter Relations}

We define a binary relation on $\List[\Fin[n]]$.

We define $\cox*$ as the congruence closure of $\cox$.
\vc{This could be simplified a bit depending on our formalisation. $\cox$ is directed, use a different symbol?}

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item $\cox$ is (locally) confluent. For every span $w_{1} \cox w \cox w_{2}$, there is a matching extended cospan
          $w_{1} \cox* v \cox* w_{2}$.
    \item $\cox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \cox* v$.
          \review{Equivalently, the type $\dsum{v:\List[\Fin[n]]}{w \cox* v}$ is contractible, and $w \cox* v$ is a
            proposition for every $w$ and $v$.}
  \end{enumerate}
\end{proposition}

>>>>

When swapping two pairs of elements in positions "far-away", it does not matter
in which order the swapping happens, that is, we can slide the wires freely.

@@

\todo{explain braid in words, give some intution on why it holds}

@@

\todo{an example/motivation from book.pdf}

This construction is called a Coxeter presentation. Writing it formally, we get
the following definition for $\Sn$. We define a binary relation $\cox$ on
$\List[\Fin[n]]$.

We define $\cox*$ as the congruence closure of $\cox$.

The choice of generators for this presentation is important. \todo{Think about this.}

\begin{itemize}
  \item It affects the difficulty of solving the word problem in $\Sn$ and
        formalising the proof of its correctness.
  \item It affects the proof of strong normalisation of reduction for the relations.
  \item It dictates which words are normal forms in this presentation of $\Sn$.
  \item It has to closely match the $\PiLang$ comobinators so that we can prove
        completeness.
  \item It dictates the choice of reversible gates in the synthesis and
        normalisation of boolean circuits which are applications that we show in
        the later sections.
\end{itemize}

We consider the generators for the Coxeter relations on $\List[\Fin[n]]$ as a
rewriting system. Throwing out reflexivity, and symmetry, we observe that the
right hand sides of the relations are strictly smaller than the left hand sides,
in terms of the lexicographical ordering on words in $\Fin[n]$. By directing the
relation from left to right, we would be able to prove a termination property
out of the box.

However, it is not at all obvious that the system would be confluent.
There are many possible critical pairs, such as
% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzQsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJcXGJyYWlkIiwxXSxbMCwyLCJcXGJyYWlkIiwxXV0=
\[\begin{tikzcd}
    && {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
    {\tau_1\tau_2\tau_1\tau_1\tau_2} &&&& {\tau_2\tau_1\tau_1\tau_2\tau_1}
    \arrow["\braid"{description}, from=1-3, to=2-1]
    \arrow["\braid"{description}, from=1-3, to=2-5]
  \end{tikzcd}\]
or
% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFs0LDEsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiXFxicmFpZCIsMV0sWzAsMiwiXFxjYW5jZWwiLDFdXQ==
\[\begin{tikzcd}
    && {\tau_2\tau_1\tau_2\tau_2} \\
    {\tau_1\tau_2\tau_1\tau_2} &&&& {\tau_2\tau_1}
    \arrow["\braid"{description}, from=1-3, to=2-1]
    \arrow["\cancel"{description}, from=1-3, to=2-5]
  \end{tikzcd}\]
Although in the cases above, critical pairs resolve, there are some examples where they do not. One of the simplest ones is the following:
% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMSwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFs0LDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJcXGJyYWlkIiwxXSxbMCwyLCJcXHN3YXAiLDFdXQ==
\[\begin{tikzcd}
    && {\tau_3\tau_2\tau_3\tau_1} \\
    {\tau_2\tau_3\tau_2\tau_1} &&&& {\tau_3\tau_2\tau_1\tau_3}
    \arrow["\braid"{description}, from=1-3, to=2-1]
    \arrow["\swap"{description}, from=1-3, to=2-5]
  \end{tikzcd}\]

Because of this counterexample, the equations have to be fixed. We thus propose a new presentation, keeping the
$\cancel$ and $\swap$ constructors, but replace $\braid$ with a more general $\longbraid$:

@@

\begin{definition}[$\longcox$]
  \begin{align*}
    \longcancel
    & : \forall n, l, r \to (l \append n \cons n \cons r) \longcox (l \append r) \\
    \longswap
    & : \forall k, n, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r) \\
    \longbraid
    & : \forall n, l, r \to (l \append (n \downarrow 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downarrow 2 + k) \append r) \\
  \end{align*}
\end{definition}

\begin{definition}[$\longcox*$]
  \begin{align*}
    \reflr{\longcox}
    & : \forall w \to w \longcox* w \\
    \transr{\longcox}
    & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3} \\
  \end{align*}
\end{definition}

\review{%
  In return for increased complexity of the equations, it satisfies the diamond property (if we direct the equation).
  Thus, together with the (preserved) termination, it creates a confluent rewriting system. Proof: All the critical
  pairs resolve, in the appendix. Definition: A normal form is a list that cannot be reduced any further.%
}

Now, we state and prove that $(\List[\Fin[n]],\longcox*)$ is a confluent and strongly normalising rewriting system.

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item $\longcox*$ is (locally) confluent. For every span $\longcoxspan{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\longcoxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is confluent. For every extended span $\longcoxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\longcoxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
    \item $\cox*$ and $\longcox*$ are equivalent. For every $w$ and $v$, $w \cox* v$ iff $w \longcox* v$.
  \end{enumerate}
\end{proposition}

@@
